Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-04-26 papers/histogram/sad.tex
these are the paper edits and fixed horizontal plotting range to be consistent for figures.

@@

\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}
\newcommand{\jpsays}[1]{{\color{red} [\blue{Jordan:} \emph{#1}]}}
\newcommand{\tssays}[1]{{\color{red} [\cyan{Tanner:} \emph{#1}]}}

\begin{document}
\title{Stochastic Approximation Monte Carlo with a Dynamic Update
Factor (SAD)
}

@@

Wang and Landau introduced an algorithm (WL) that uses an update
factor and a statistical histogram to compute the density of states of
a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has a few disadvantage.
Firstly, it requires the user to select a number of parameters,
including the range of energies to be studied~\cite{landau2004new}.
This adds an additional hurdle to its application to a variety of
systems.  The simulation violates detailed balance, with the size of
the violation decreasing with time, which complicates convergence
analysis.  In fact, the error in a WL computation the error has been
demonstrated to saturate at a non-zero value~\cite{yan2003fast},
i.e. the method does not converge to the true density of
states~\cite{belardinelli2007wang, belardinelli2007fast,
  belardinelli2008analysis, belardinelli2014intrinsic,
  singh2012density}.  
%
Belardinelli and Pereyra demonstrated that allowing the update factor
to decrease faster than $1/t$ leads to
nonconvergence~\cite{belardinelli2007wang}.
%% Around the same time, the convergence rate for WL was being
%% explored~\cite{zhou2005understanding,lee2006convergence,
%%   belardinelli2007wang}.
%
This leads to their $1/t$-WL algorithm which ensures that the error
continues to decrease asymptotically as
$1/\sqrt{t}$~\cite{belardinelli2008analysis}.

Liang, Liu, and Carrol began to consider whether WL could be considered
a special case of Stochastic Approximation whose convergence could be
mathematically proven~\cite{liang2006theory, liang2007stochastic}. In
2007, Liang et al.~\cite{liang2007stochastic} argued that WL can be
considered a form of Stochastic Approximation Monte Carlo (SAMC). While
SAMC can guarantee convergence, the method still has a system specific
user-defined variable which makes applying this algorithm to arbitrary
systems difficult.

In this work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Stochastic Approximation
with a dynamic update factor), and will discuss it in detail in the methods
section. We compare it along with several flat histogram methods
which include WL, $1/t$-WL, and SAMC.

@@

The organization of this paper is as follows: In Section~\ref{sec:histogram}, we
describe in detail flat-histogram methods.
%
Section~\ref{sec:sad} introduces the dynamic stochastic approximation method.
In Section~\ref{sec:results}, we discuss the performance of the various methods
applied to the model system.

In this work, We compare a variety of flat histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are introduced and applied to the
square-well fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte-Carlo (SAMC), and Stochastic Approximation with a
dynamic $\gamma$ (SAD).

@@

The goal of flat histogram methods (also called \emph{broad histogram}
or \emph{multicanonical} methods) is to simulate each energy with
similar accuracy, so as to accurately determine the density of states
over a broad range of energies---and thus to determine the thermodynamic
quantities such as heat capacity or internal energy
over a broad range of temperatures.
Properties that require more information---such as a spatial
correlation function or a response function---can still be computed
for any temperature, provided statistics are collected for each
individual energy, which can then be reweighted for any temperature.

Each flat histogram Monte Carlo method uses a given set of ``moves''
which change a system and should satisfy detailed balance.  Each
algorithm differs in how it determines the probability of accepting a
move, and in what additional statistics must be collected in
order decide on that probability.

@@

Each approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma
\end{equation}
where $t$ is number of the current move, and $\gamma$ is an update
factor.  This update causes the random walk to avoid energies that
have been frequently sampled, leading to a rapid exploration of energy
space.  This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreased as we decrease $\gamma$.  The various
weight-based methods discussed here differ in how the decreasing of $\gamma$
is scheduled.

@@

While this approach
is very efficient and has been widely used, it suffers a few
shortcomings.  Firstly, the set of energies must be specified
~\cite{wang2001efficient, schulz2003avoiding, yan2003fast}, which may
require multiple simulations. Secondly, while Wang-Landau converges quickly,
it does not in general converge to the true density of states
~\cite{belardinelli2008analysis, zhou2008optimal}.  It can and does decrease
$\gamma$ so quickly that it will never (for any cutoff value) decrease the
error in the density of states beyond a given value.

\subsection{$1/t$-Wang-Landau}
The Wang-Landau algorithm's error saturation can be corrected by
modifying the update factor such that it does not decrease too quickly.
Belardinelli and Pereyra raised the question as to whether the update
factor should be decrease by $\nicefrac12$ or some other factor for
each update~\cite{belardinelli2007fast}. They implemented a schedule
that enforced if the update factor $\gamma$ is less than
$\nicefrac{N_S}{t}$ then the update factor is set to
$\nicefrac{N_S}{t}$ and the histogram is no longer tracked. Employing
this scheduler, they found that the error saturation is avoided since
the correct density of states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between
$t^{-\frac12}$ and $1/t$~\cite{zhou2008optimal}.

@@

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$.
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameter $t_0$.  Unfortunately, there is no
prescription for finding an acceptable value for $t_0$, and
while the algorithm formally converges, for a poor choice of $t_0$
that convergence can be far too slow to be practical.
Liang et al. give a rule of thumb in which $t_0$
is chosen in the range from $2N_S$ to $100N_S$ where $N_S$ is the number
of energy bins~\cite{liang2007stochastic}.  Schneider et al. found
that for the
Ising model this heuristic is helpful for small spin
systems, but that larger systems require an even higher $t_0$
value~\cite{schneider2017convergence}.  Also, we find that for the square-well
fluid $t_0$ in some cases needs to be two orders of magnitude higher
than the rule of thumb of $100N_S$.

Werlich \emph{et al.} proposed scaling the SAMC $\gamma_t$ by a factor
$\gamma_0$.  While this may result in an improved rate of convergence,
it adds yet another parameter that must be empirically determined, and
we have not explored this degree of freedom.

\subsection{SAMC Convergence}\label{sec:samc-convergence}
The difficulty in using the SAMC method lies in identifying an appropriate
value for $t_0$.  A value that is either too high or too low will result
in slow convergence to the true entropy.  It is instructive to consider
separately values of $t_0$ that are too low or too high.

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true energy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
The minimum number of moves that could converge the entropy is
the number of moves that will enable the above total entropy change,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t \\
  % &= \int_0^{t_{\min}} \gamma(t)dt \\
  &= t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to convergence grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

One might reasonably choose to err on the high side when selecting a $t_0$.
The rate of convergence is harder to estimate when $t_0$ is large, but
in general $\gamma_t$ itself forms a lower bound on the accuracy with which
the entropy may be known, with an unknown prefactor which is roughly
the coherence time of the Monte Carlo simulation.  Since $\gamma_t$ is
given by $t_0/t$, the time to converge to a given accuracy is increased
in proportion to the ratio by which we overestimate $t_0$.  Thus, while
it is exponentially painful to underestimate $t_0$, overestimating by
several orders of magnitude is also not acceptable.  We should note
that these extreme limiting cases do not preclude the possibility that there
is a wide range of $t_0$ values that lead to an acceptable convergence
rate.

@@

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by
$E_H -E_L$ where $E_H$ and $E_L$ are defined below.
The entropy numerator of the update factor
in general should scale with the total number of interesting
energy states $N_S$, since updates to the weights are distributed between
that many energy states.  The product $N_S\langle S\rangle$ is the
total change of entropy required (starting from constant weights) to find
the true entropy, and puts a lower bound on the convergence time.
After \emph{long} times, when
all the energies have been long ago found,
we wish for a lower update factor in order to more rapidly refine the remaining
error in entropy.  We track the time at which we first visted each possible
energy.  We define $t_L$ to be the last time that we encountered an energy
that we currently believe is in the energy range of interest, so a $t\gg t_L$
we feel confident that we have established the true energy range of interest. We
gradually transition to a lower update factor (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  Finally,
we wish for an update factor that is \emph{never} greater than 1, because
a very large update factor could introduce errors in entropy that may take
many iterations to remove.  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

@@

Since SAD does not explore \emph{all} energy states it needs determine
what energy range corresponds to the temperature range of interest,
defined by $T_{\min}<T<\infty$.
The
simulation is responsible for determining and updating this energy
range.
Given the true entropy $S(E)$, we can define the interesting energy
range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  In the course of the
simulation this precise energy is challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$, which define the range over which the energy histogram
is made flat. At move $t$, $E_H$ and $E_L$ are the greatest and lowest
energy that has prior to that move
had the highest histogram value (i.e. been visited the most times) in
the course of the simulation.
This definition results in a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

During the simulation when considering a move outside of the energy
range $E_L \le E \le E_H$, Boltzmann weights are used.  If $E\ge E_H$,
the weight is taken to be
\begin{align}\label{eq:highw}
  w(E>E_H) &= w(E_H),
\end{align}
which corresponds to an infinite temperature.
This choice ensures that
if the maximum in entropy is at an energy $E_{\max}>E_H$, then the energy
$E_{\max}$
will eventually have the highest number of counts and the ratcheting will
result in $E_H\ge E_{\max}$ .
At lower energies, Boltzmann weights corresponding to the minimum temperature
are used:
\begin{align}\label{eq:loww}
  w(E<E_L) &= w(E_L)e^{-\frac{E_L-E}{T_{\min}}}.
\end{align}
This choice has the result that if the energy $E_{\min}$ at which the free
energy at $T_{\min}$ is minimized is less than $E_L$, the lower energy
limit will ratchet down to include $E_{\min}$.
Each time we change the value of $E_H$ or $E_L$, the weights within the
new portion of the interesting energy range are set to the expressions
in Equations~\ref{eq:highw} and~\ref{eq:loww}.

@@

A significant advantage of SAD over other flat histogram methods is
that energy binning is handled dynamically. SAD should perform
independently for a reasonable range of binning because $\gamma \propto
\nicefrac{N_S}{t}$.  As the number of energy states found $N_S$
increases (fine binning), the time spent $t$ in each bin will decrease
with the effect that the convergence should be roughly independent of
the bin size chosen.
This argument explains why SAD should perform well for continuum systems
in terms of binning. SAMC has a prefactor $\gamma_0$ to aid in a
similar way but this adds yet another parameter for the user to choose.

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run.  The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area, and
           (b) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
           (c) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.005\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run, and
           (d) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
  }\label{fig:fast-slow-gamma}
\end{figure*}

@@

For this simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from $-248$ to $-120$.  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198, corresponding to a ratio of
$10^{86}$ between the highest and lowest density of states.

In order to explore the effect of simulation details on convergence,
we consider two values for the mean distance by which atoms are moved
during a Monte Carlo step.  We began with a reasonable displacement
distance of $\delta_0 = 0.05\sigma$, which corresponds to an
acceptance rate of 38\%.  We further ran simulations with a much
smaller displacement distance of $\delta_0 = 0.005\sigma$, which
resulted in an acceptance rate of 86\%, which we expected converge
more slowly.

@@

Figure~\ref{fig:fast-slow-gamma}a shows the average error in the
entropy as a function of time for this system with the reasonable
displacement distance of $\delta_0 = 0.05\sigma$.  The solid lines
represent the average of the absolute value of the error in the
entropy averaged over eight simulations using different random number
seeds.  The range of average errors for each simulation is shown as a
shaded region around its mean error.  By the time $10^8$ moves have
been made all but the SAMC simulation with the shortest $t_0$ have
begun to converge as $1/\sqrt{t}$.  We then see the WL error begin to
saturate around $10^{10}$ moves.

>>>>

\usepackage{xargs}                      % Use more than one optional parameter in a new commands
\usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.
\usepackage[colorinlistoftodos,prependcaption,textsize=normalsize]{todonotes}
\usepackage{mdframed}

% define colors for comments
\definecolor{dark-gray}{gray}{0.10}
\definecolor{light-gray}{gray}{0.70}

@@

\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}
\newcommand{\jpsays}[1]{{\color{red} [\blue{Jordan:} \emph{#1}]}}
\newcommandx{\jpcom}[2][1=inline]{\todo[linecolor=gray,backgroundcolor=light-gray,bordercolor=dark-gray,#1]{\textbf{Jordan says:} #2} }
\begin{document}

\title{Stochastic Approximation Monte Carlo with a Dynamic Update
Factor (SAD)
}

@@

Wang and Landau introduced an algorithm (WL) that uses an update factor
and a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has a few disadvantages. Firstly, it
requires the user to select a number of parameters, including the range
of energies to be studied~\cite{landau2004new}. This adds an additional
hurdle to its application to a variety of systems because the user must
simulate until the energy range of interest is known.  The simulation
violates detailed balance albiet briefly as the size of the violation decreases
with time, which complicates convergence analysis.  In fact, the error
in a WL computation the error has been demonstrated to saturate at a
non-zero value~\cite{yan2003fast}, i.e. the method does not converge to
the true density of states~\cite{belardinelli2007wang,
belardinelli2007fast, belardinelli2008analysis,
belardinelli2014intrinsic, singh2012density}.
%
Belardinelli and Pereyra demonstrated that allowing the update factor
to decrease faster than $1/t$ leads to
nonconvergence~\cite{belardinelli2007wang}.
%% Around the same time, the convergence rate for WL was being
%% explored~\cite{zhou2005understanding,lee2006convergence,
%%   belardinelli2007wang}.
%
This leads to their $1/t$-WL algorithm which ensures that the error
continues to decrease asymptotically as
$1/\sqrt{t}$~\cite{belardinelli2008analysis}.

Liang, Liu, and Carrol began to consider that WL could be considered
a special case of Stochastic Approximation whose convergence could be
mathematically proven~\cite{liang2006theory, liang2007stochastic}. In
2007, Liang et al.~\cite{liang2007stochastic} argued that WL can be
considered a form of Stochastic Approximation Monte Carlo (SAMC). While
SAMC can guarantee convergence, the method still has a system specific
user-defined variable which makes applying this algorithm to arbitrary
systems difficult.

In this work, we have developed a novel algorithm based on SAMC that
does not require an array of non-physical, user-defined inputs and
therefore should be easily applicable to a given system. The method
does require the user to define a temperature range ($T_\text{min}$ to $T_\infty$).
We call this method SAD (Stochastic Approximation with a dynamic update
factor), and will discuss it in detail in the methods section. We
compare it along with several flat histogram methods which include WL,
$1/t$-WL, and SAMC.

@@

%The organization of this paper is as follows: In Section~\ref{sec:histogram}, we
%describe in detail flat-histogram methods.
%Section~\ref{sec:sad} introduces the dynamic stochastic approximation method.
%In Section~\ref{sec:results}, we discuss the performance of the various methods
%applied to the model system.

In this work, We compare a variety of flat histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are discussed and simulated for the
square-well fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte-Carlo (SAMC), and Stochastic Approximation with a
dynamic $\gamma$ (SAD).

@@

The goal of flat histogram methods (also called \emph{broad histogram}
or \emph{multicanonical} methods) is to simulate each energy with
similar accuracy, so as to accurately determine the density of states
over a broad range of energies---and thus to determine the thermodynamic
quantities such as heat capacity or internal energy
over a broad range of temperatures.
Properties that require more information---such as a spatial
correlation function or a response function---can still be computed
for any temperature, provided statistics are collected for each
individual energy, which can then be reweighted for any
temperature~\cite{errington2003direct}.

Each flat histogram Monte Carlo method uses a given set of ``moves''
which change the system and should satisfy detailed balance.  Each
algorithm differs in how it determines the probability of accepting a
move and in what additional statistics must be collected in
order decide on that probability.

@@

Each approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma
\end{equation}
where $t$ is number of the current move, and $\gamma$ is an update
factor.  This update causes the random walk to avoid energies that
have been frequently sampled, leading to a rapid exploration of energy
space.  This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreases as we decrease $\gamma$.  The various
weight-based methods discussed here differ in how they schedule the
decreasing of $\gamma$.

@@

While this approach
is very efficient and has been widely used, it suffers a few
shortcomings.  Firstly, the set of energies must be specified
~\cite{wang2001efficient, schulz2003avoiding, yan2003fast}, which may
require multiple simulations. Secondly, while Wang-Landau converges quickly,
it does not in general converge to the true density of states
~\cite{belardinelli2008analysis, zhou2008optimal}.  It can and does decrease
$\gamma$ so quickly that it will never (for any cutoff value) decrease the
error in the density of states beyond a given nonzero value.

\subsection{$1/t$-Wang-Landau}
The Wang-Landau algorithm's error saturation can be corrected by
modifying the update factor such that it does not decrease too quickly.
Belardinelli and Pereyra raised the question as to whether the update
factor should be decrease by $\nicefrac12$ or some other factor for
each update~\cite{belardinelli2007fast}. They implemented a schedule
that enforced that if the update factor $\gamma$ is less than
$\nicefrac{N_S}{t}$ then the update factor is set to
$\nicefrac{N_S}{t}$ where $N_S$ is the number of energy states and the
histogram is no longer tracked. Employing this scheduler, they found
that the error saturation is avoided since the correct density of
states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between $t^{-\frac12}$ and
$1/t$~\cite{zhou2008optimal}.

@@

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
The major advantage that SAMC offers is its proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
liang2007stochastic}.  In addition, the energy range need not be known
a priori.  The time to converge depends only on the choice of parameter
$t_0$.  Unfortunately, there is no prescription for finding an
acceptable value for $t_0$, and while the algorithm formally converges,
for a poor choice of $t_0$ that convergence can be far too slow to be
practical. Liang et al. give a rule of thumb in which $t_0$ is chosen
in the range from $2N_S$ to $100N_S$ where $N_S$ is the number of
energy bins~\cite{liang2007stochastic}.  Schneider et al. found that
for the Ising model this heuristic is helpful for small spin systems,
but that larger systems require an even higher $t_0$
value~\cite{schneider2017convergence}.  Also we find that for one case
with a small displacement distance for the square-well fluid, $t_0$
needs to be as much as two orders of magnitude higher than the rule of
thumb of $100N_S$ in order to converge in $10^{12}$ moves.

Werlich \emph{et al.} proposed scaling the SAMC $\gamma_t$ by a factor
$\gamma_0$~\cite{werlich2015stochastic}.  While this may result in an
improved rate of convergence, it adds yet another parameter that must
be empirically determined, and we have not explored this degree of
freedom.

\subsection{SAMC Convergence}\label{sec:samc-convergence}
A primary difficulty in using the SAMC method lies in identifying an appropriate
value for $t_0$.  A value that is either too high or too low will result
in slow convergence to the true entropy of the system.  It is instructive to consider
separately values of $t_0$ that are too low or too high.

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true entropy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
The minimum number of moves that could converge the entropy is
the number of moves that will enable the above total entropy change,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t \\
  % &= \int_0^{t_{\min}} \gamma(t)dt \\
  &= t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to convergence grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

One might reasonably choose to err by selecting a large $t_0$.
The rate of convergence is harder to estimate when $t_0$ is large, but
in general $\gamma_t$ itself forms a lower bound on the accuracy with which
the entropy may be known, with an unknown prefactor which is roughly
the coherence time of the Monte Carlo simulation.  Since $\gamma_t$ is
given by $t_0/t$, the time to converge to a given accuracy is increased
in proportion to the ratio by which we overestimate $t_0$.  Thus, while
it is exponentially painful to underestimate $t_0$, overestimating by
several orders of magnitude is also not acceptable.  We should note
that these extreme limiting cases do not preclude the possibility that there
is a wide range of $t_0$ values that lead to an acceptable convergence
rate.

@@

While for SAMC, the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by $E_H -E_L$ where $E_H$ and
$E_L$ are defined below. The entropy numerator of the update factor in
general should scale with the total number of interesting energy states
$N_S$, since updates to the weights are distributed between that many
energy states.  The product $N_S\langle S\rangle$ is the total change
of entropy required (starting from constant weights) to find the true
entropy, and puts a lower bound on the convergence time. After
\emph{long} times, when all the energies have been a long time ago, we
wish for a lower update factor in order to more rapidly refine the
remaining error in entropy.  We track the time at which we first visted
each possible energy.  We define $t_L$ to be the last time that we
encountered an energy that we currently believe is in the energy range
of interest, so a $t\gg t_L$ we feel confident that we have established
the true energy range of interest. We gradually transition to a lower
update factor (but still asymptotically scaling as $\gamma_t \propto
1/t$ to ensure eventual convergence).  Finally, we wish for an update
factor that is \emph{never} greater than 1, because a very large update
factor could introduce errors in entropy that may take many iterations
to remove.  The SAD expression for $\gamma_t$ which incorporates these
ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

@@

Since SAD does not explore \emph{all} energy states, it needs to
determine what energy range corresponds to the temperature range of
interest defined by $T_{\min}<T<\infty$. The simulation is responsible
for determining and updating this energy range. Given the true entropy
$S(E)$, we can define the interesting energy range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  During the course of the
simulation, this precise energy is challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$, which define the range over which the energy histogram
is made flat. At move $t$, $E_H$ and $E_L$ are the greatest and lowest
energy that prior to that move
had the highest histogram value (i.e. been visited the most times) during
the course of the simulation.
This definition results in a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

During the simulation when considering a move inside of the energy
range $E_L \le E \le E_H$, Boltzmann weights are used.  If $E\ge E_H$,
the weight is taken to be
\begin{align}\label{eq:highw}
  w(E>E_H) &= w(E_H),
\end{align}
which corresponds to an infinite temperature.
This choice ensures that
if the maximum in entropy is at an energy $E_{\max}>E_H$, then the energy
$E_{\max}$
will eventually have the highest number of counts and the ratcheting will
result in $E_H\ge E_{\max}$ .
At lower energies, Boltzmann weights corresponding to the minimum temperature
are used:
\begin{align}\label{eq:loww}
  w(E<E_L) &= w(E_L)e^{-\frac{E_L-E}{T_{\min}}}.
\end{align}
This choice has the result that if the energy $E_{\min}$ at which the free
energy at $T_{\min}$ is minimized is less than $E_L$, the lower energy
limit will ratchet down to include $E_{\min}$.
Each time we change the value of $E_H$ or $E_L$, the weights within the
new portion of the interesting energy range are set to the expressions
in Equations~\ref{eq:highw} and~\ref{eq:loww}.

@@

A significant advantage of SAD over other flat histogram methods is
that energy binning is handled dynamically. SAD should perform
independently for a reasonable range of binning because $\gamma \propto
\nicefrac{N_S}{t}$.  As the number of energy states $N_S$ found
increases (fine binning), the time spent $t$ in each bin will decrease
with the effect that the convergence should be roughly independent of
the bin size chosen.
This argument explains why SAD should perform well for continuum systems
in terms of binning. SAMC has a prefactor $\gamma_0$ to aid in a
similar way~\cite{werlich2015stochastic} but this adds yet another parameter for the user to choose.

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run. The number of important
               energies is $129$ of about $280$ energies.  The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area, and
           (b) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
           (c) The average entropy error for each MC method for $N=50$,
               $\delta_0 = 0.005\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run. The number of important
               energies is $407$ of about $1800$ energies, and
           (d) the update factor $\gamma_t$ versus iteration number
               for the same simulations.
  }\label{fig:fast-slow-gamma}
\end{figure*}

@@

For this simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from $-248$ to $-120$.  The number of important energy states
for this system is therefore $N_S = 129$  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198, corresponding to a ratio of
$10^{86}$ between the highest and lowest density of states.

In order to explore the effect of simulation details on convergence,
we consider two values for the displacement distance by which atoms are moved
during a Monte Carlo step.  We began with a reasonable displacement
distance of $\delta_0 = 0.05\sigma$, which corresponds to an
acceptance rate of 38\%.  We further ran simulations with a much
smaller displacement distance of $\delta_0 = 0.005\sigma$, which
resulted in an acceptance rate of 86\%, which as expected converge
more slowly.

@@

Figure~\ref{fig:fast-slow-gamma}a shows the average error in the
entropy as a function of time for this system with the reasonable
displacement distance of $\delta_0 = 0.05\sigma$.  The solid/dashed lines
represent the average of the absolute value of the error in the
entropy averaged over eight simulations using different random number
seeds.  The range of average errors for each simulation is shown as a
shaded region around its mean error.  By the time $10^8$ moves have
been made all but the SAMC simulation with the shortest $t_0$ have
begun to converge as $1/\sqrt{t}$.  We then see the WL error
saturate around $10^{10}$ moves.
